NumLin: Linear Types for Linear Algebra

### 2019-03-13 write-up/response.txt

# Questions:

## Questions common to all the reviews, and big questions

## Reviewer-specific questions

### Review A

* The exact research contribution of this paper is, however, unclear
  to me. Linear types as well as fractional permissions have been used
  before.  Is the contibution that these two type systems are applied
  together, or that they are applied to a new domain? If it is the
  latter, it would be helpful if the paper could point out differences
  to previous type systems, or challenges, if there are any.

* I have an additional question: how were the types of the BLAS/LAPACK
  routines obtained? Did you just annotate the APIs based on
  assumptions documented somewhere, or where these APIs also
  type-checked?

* Finally, the related work focuses almost exclusively on type
  systems. Since ultimately the goal is to provide (memory) safe and
  fast linear algebra, I think it would be nice to also mention
  non-type based efforts, e.g. Spiral in Scala which compiles to
  efficient code [1], even though it does not use BLAS/LAPACK.  [1]
  Spiral in Scala: Towards the Systematic Construction of Generators
  for Performance Libraries. GPCE ’13

### Review B

* One question I have is why OCaml code is generated? If the program
  only uses higher-order functions in certain ways, would it be
  possible to generate C code?

* Also, could the aliasing requirements be checked dynamically? I.e.,
  provide the OCaml interface in Appendix G, but with dynamic aliasing
  checks? What overhead would that have?

* I was confused by the desugaring "let !x = e1 in e2" => `let Many x
  = e1 in let Many x = Many (Many x) in e2`. Can you explain this
  more?

* The statement to line 302 talks about moving values from the linear
  environment to the intuitionistic one. But only variables live in
  environments?

* I was initially confused by the presence (in Figure 11) of the
  seemling useless wrappers around banged values. Is this to make the
  API of the generated OCaml code easier to use correctly?

* I was confused by the comment on Linear Haskell starting on line
  610. What relevance does a system based on kinds have to NumLin
  (which isn't based on kinds either)?

### Review C

* How exactly could they incorporate this knowledge of what is
  legal/illegal wrt Fortran in the Numlin framework? Or was it spotted
  incidentally when they examined the permissions under which an
  external routine was attempted to be invoked?

* Compared to the learning curve of Rust (which the authors cite) what
  is the learning curve of Numlin, in the authors' estimation?

* Could the authors sketch the evolution of linear types from its
  initial classical introduction (say in Wadler's "Linear types can
  change the world" paper) to its current sophisticated form? Is this
  really an "off the shelf" linear type system?

* Can they explain their soundness theorem once in English?

* My biggest comment is concerning concurrency. These sort of routines
  will be used in a setting where multicore concurrency will be
  employed to perform concurrent updates on arrays. In that setting,
  one might employ separation logic to deal with safe concurrent
  accesses. Now, separation logic also employs fractional
  permissions. Would Numlin be usable in that setting? Would the use
  of Separation Logic and Numlin (Linear Types) be orthogonal? Or do
  the authors feel the need to work out a concurrency extension of
  Numlin?

----

ECOOP '19 Paper #35 Reviews and Comments
===========================================================================
Paper #35 NumLin: Linear Types for Linear Algebra

Review #35A
===========================================================================

Overall merit
-------------
C. Weak reject: I think this paper should be rejected but I will not argue
   strongly against it

Reviewer expertise
------------------
Y. Knowledgeable

Paper summary
-------------
The paper presents a language and type system called NumLin for writing linear algebra programs which use BLAS/LAPACK routines.  NumLin's type system is based on linear and fractional permission types and ensures correct memory usage and aliasing. The performance achieved is comparable to an unsafe C implementation.

Comments for author
-------------------
The paper targets an important domain; high-performance linear algebra routines are used in many different applications.  However, common libraries such as BLAS or LAPACK are written in low-level languages and require careful memory management. Providing more (memory) safety without sacrificing performance is thus of practical importance.  
The paper demonstrates how to provide this safety using a type system, and provides a prototype solution in the form of a  domain-specific language which compiles down to OCaml. Thus, it provides a practical implementation which is of potential interest to a wide audience.

The exact research contribution of this paper is, however, unclear to me. Linear types as well as fractional permissions have been used before.  Is the contibution that these two type systems are applied together, or that they are applied to a new domain? If it is the latter, it would be helpful if the paper could point out differences to previous type systems, or challenges, if there are any.

My lack of understanding of the contribution is probably also due to the fact that I found the paper hard to read:
- For instance, the exact usage of the language was unclear to me initiatially. The beginning of the paper talks about expressing the APIs of BLAS/LAPACK libraries, but much what follows is about writing programs (using those APIs), but the APIs are hidden. 
- Similarly, the overview of NumLin in section 2 read like a background section, but did not really explain to me how this fits together in NumLin.
- The examples in section 2.2 hinted at different features, but left too many details for later to help understanding. I found the examples were trying to introduce too much at the same time: syntax of the language, syntactic sugar, different BLAS routines which are supported, as well as the type system. For me, it would have been clearer to discuss these (more) individually, and I believe this would make the paper more accessible to a wider audience.

I have an additional question: how were the types of the BLAS/LAPACK routines obtained? Did you just annotate the APIs based on assumptions documented somewhere, or where these APIs also type-checked?

Finally, the related work focuses almost exclusively on type systems. Since ultimately the goal is to provide (memory) safe and fast linear algebra, I think it would be nice to also mention non-type based efforts, e.g. Spiral in Scala which compiles to efficient code [1], even though it does not use BLAS/LAPACK.
[1] Spiral in Scala: Towards the Systematic Construction of Generators for Performance Libraries. GPCE ’13

Review #35B
===========================================================================

Overall merit
-------------
A. Accept: I will champion this paper

Reviewer expertise
------------------
X. Expert

Paper summary
-------------
The APIs of low level linear algebra libraries, like LAPACK and BLAS go to great lengths to allow users full flexibility in how memory to store matricies is allocated, freed, and reused. However, this comes at the cost of potential memory safety bugs as the API does not automatically enforce the (non-)aliasing assumed by the implementations of the routines. High-level wrappers around these APIs ensure memory safety, but at the cost of additional temporary storage and defensive copying.

This paper proposes a "middle way": a high level programming language that directly uses the low-level routines, and uses a linear type system to ensure that the (non-)aliasing constraints required by the low-level routines are satisfied. The system is proved sound using a step-indexed logical relation. An implementation has been constructed that generates OCaml code, and it has been benchmarked against high-level code in Owl (an OCaml library) and NumPy (a Python library). The benchmark results show a convincing speedup in cases where the execution time is not dominated by the low level routines themselves.

Comments for author
-------------------
This paper introduces a non-trivial system to solve a real problem, provides both a proof of soundness and an implementation, and uses that implementation to run a convincing evaluation. I recommend this paper for acceptance.

One question I have is why OCaml code is generated? If the program only uses higher-order functions in certain ways, would it be possible to generate C code?

Also, could the aliasing requirements be checked dynamically? I.e., provide the OCaml interface in Appendix G, but with dynamic aliasing checks? What overhead would that have?

Points in favour of this paper:

  - The type system is introduced in a clear and easy to follow way, with the linear aspects of the type system introduced in a logical order. Several examples are given, progressing from fibonacci to relatively involved examples, showing that the library is actually usable for real work.

  - There is a soundness proof, carried out using step-indexed logical relations. While this is an adaptation of previous work by Morrisett et al., it is non-trivial to apply it to this system, and it includes the novel feature (for a type system like this) of using fractional permissions.

  - The inclusion of fractional permissions into a unification based type inference system appears to be novel, and while the paper does take pains to not claim that this feature has been formally proved complete, it does make the implementation feasible to use for real examples.

  - The existence of an implementation means that a good empirical evaluation can be carried out, and this shows that the NumLin system does indeed have the claimed benefits of reducing the overhead of generating temporary arrays and low-level copying.

Things that could be improved / suggestions:

  - The type system appears to be based on Andrew Barber's DILL (Dual Intuitionistic Linear Logic), but this is not referenced.

  - I was confused by the desugaring "let !x = e1 in e2" => `let Many x = e1 in let Many x = Many (Many x) in e2`. Can you explain this more?

  - The statement to line 302 talks about moving values from the linear environment to the intuitionistic one. But only variables live in environments?

  - The "value" restriction in the !-introduction rule should be explained more, with an example that shows a "bad" use. I guess the restriction arises from the existence of allocating primtivies in the language, such as `array`, and allowing these in the !-introduction rule would permit unrestricted duplication of references to newly allocated elements (another way round this would be to have a token passed around encapsulating the ability to allocate). Ahmed et al. (2007) [1] also have a similar value restriction for the same reason (I also think this would be a better reference than the Morrisett et al. TLCA2005 paper).

  - The types of some of the primitive should be given in the paper's main body, so that it is possible to see how to type the aliasing restrictions required for a given routine. As given, the reader has to reverse engineer the NumLin type from the OCaml type given in the appendix.

  - I was initially confused by the presence (in Figure 11) of the seemling useless wrappers around banged values. Is this to make the API of the generated OCaml code easier to use correctly?

  - I was confused by the comment on Linear Haskell starting on line 610. What relevance does a system based on kinds have to NumLin (which isn't based on kinds either)?

Typos:

  - Line 327: "follow" => "follows"

  - Line 494: Missing words at the start of the line.

[1] Amal Ahmed, Matthew Fluet, Greg Morrisett: L3: A Linear Language with Locations. Fundam. Inform. 77(4): 397-449 (2007)

Review #35C
===========================================================================

Overall merit
-------------
A. Accept: I will champion this paper

Reviewer expertise
------------------
Z. Outsider

Paper summary
-------------
This paper is on the use of linear types in a new programming language NUMLIN to express a large array of important algorithms involving aggregate data types (arrays primarily). Existing array based programs end up introducing many errors that are hard to detect including incorrect aliases. There also are sources of inefficiency including useless temporaries. The use of linear types that these authors propose elegantly characterizes the usage disciplines intended and also helps catch these bugs.

Comments for author
-------------------
While I am an outsider, I really find the use of linear types to be quite fascinating in terms of how it has advanced since the 80s and now applies to practical routines. The authors do a fantastic job covering a whole gamut of challenging case studies.

NUMLIN appears to impose some modest burden on part of the programmer to "correctly" annotate a program as to the intended uses of the underlying array types. Given this, the typing system can infer permissions and check for correct usage of aggregate data structures. For instance, "free" or "set" need full permissions. Multiple read uses of an array are supported via fractional permissions.

In terms of results, the slowdown is not that pronounced even compared to a C implementation. They also detected an incorrect usage in the setting of the Kalman filter wrt Fortran conventions.

Here are my suggestions and questions:

1) How exactly could they incorporate this knowledge of what is legal/illegal wrt Fortran in the Numlin framework? Or was it spotted incidentally when they examined the permissions under which an external routine was attempted to be invoked?

2) Page 2: achieving parity with C -> explain this as parity wrt runtime efficiency (no other parity with C, I suppose)

3) Compared to the learning curve of Rust (which the authors cite) what is the learning curve of Numlin, in the authors' estimation?

4) Could the authors sketch the evolution of linear types from its initial classical introduction (say in Wadler's "Linear types can change the world" paper) to its current sophisticated form? Is this really an "off the shelf" linear type system? Giving a history of the evolution of linear types to the kind used in Numlin would be very valuable

5) Can they explain their soundness theorem once in English? This can help novices read the paper better

6) My biggest comment is concerning concurrency. These sort of routines will be used in a setting where multicore concurrency will be employed to perform concurrent updates on arrays. In that setting, one might employ separation logic to deal with safe concurrent accesses. Now, separation logic also employs fractional permissions. Would Numlin be usable in that setting? Would the use of Separation Logic and Numlin (Linear Types) be orthogonal? Or do the authors feel the need to work out a concurrency extension of Numlin? This would be highly relevant to discuss, as otherwise it is unclear how we are going to employ multicores on these class of routines (run on a single core, the performance would be terrible).

### 2019-03-13 write-up/response.txt

# Questions:

## Questions common to multiple reviews

1. Reviews A and C both ask how we got the Numlin types for the
   BLAS/LAPACK routines.

The documentation records the expected aliasing behaviour of each of
these routines, and whether or not the routine modifies or consumes
its argument in any way. We used that to derive the types we ascribed
to the low-level routines. Since most of these low-level routines are
very careful not to do any allocation themselves, it was generally
very easy to give them a Numlin type -- every argument that can
modify/consume its argument needs a full permission, and all others
can be fraction-polymorphic.

@@

## Reviewer-specific questions

### Review A

* The exact research contribution of this paper is, however, unclear
  to me. Linear types as well as fractional permissions have been used
  before.  Is the contibution that these two type systems are applied
  together, or that they are applied to a new domain? If it is the
  latter, it would be helpful if the paper could point out differences
  to previous type systems, or challenges, if there are any.

We certainly agree that both linear types and fractional permissions
have been studied before. To our surprise, it did not appear that
linear types had been much applied to checking linear algebra
libraries. It was only when we tried to use it, that we realized that
we needed fractional permissions.

In our view, this explains why it hadn't been used before: many prior
type systems for fractional permissions (eg, from Boyland's original
work to Aldrich and his students' work on Plural) are extremely
complex. This is because these type systems typically encode a complex
analysis to automatically infer how fractional permissions should be
split and rejoined.

By requiring sharing and merging to be made explicit, we were able to
drastically simplify the type system. Our formal system is very close
to standard presentations of linear logic, and the implementation
complexity is no worse than HM type inference.

@@

* Finally, the related work focuses almost exclusively on type
  systems. Since ultimately the goal is to provide (memory) safe and
  fast linear algebra, I think it would be nice to also mention
  non-type based efforts, e.g. Spiral in Scala which compiles to
  efficient code [1], even though it does not use BLAS/LAPACK.  [1]
  Spiral in Scala: Towards the Systematic Construction of Generators
  for Performance Libraries. GPCE ’13

We're happy to discuss this in the related work. However, NumLin is
more likely to be a consumer of efforts like Spiral than a competitor:
we want to check whether client code conforms to the aliasing rules
that low-level libraries (whether from platform libraries like BLAS or
generated via systems like Spiral) want clients to adhere to.

@@

### Review B

* One question I have is why OCaml code is generated? If the program
  only uses higher-order functions in certain ways, would it be
  possible to generate C code?

There's no technical obstacle to generating C code, but as we had
local access to the developer of the Owl numerical programming library
for Ocaml, it made sense to use Ocaml. Eventually we'd like to extend
Owl with something like Numlin to offer safe access to the low-level
platform numerics libraries.

@@

* Also, could the aliasing requirements be checked dynamically? I.e.,
  provide the OCaml interface in Appendix G, but with dynamic aliasing
  checks? What overhead would that have?

It might be possible, but we're not sure how. Our system enforces full
linearity (to support explicit deallocation), and to our knowledge
contract systems for linearity checking (eg, Tov and Pucella's ESOP
2010 "Stateful Contracts for Affine Types") only enforce affine usage
(ie, a dynamic check can enforce at-most-one, but not exactly-once).

@@

* I was confused by the desugaring "let !x = e1 in e2" => `let Many x
  = e1 in let Many x = Many (Many x) in e2`. Can you explain this
  more?

[TODO]

Yes, this is correct. Ocaml's underlying value representation does not
allocate anything for unary datatypes, so there is no performance
penalty for using them like Haskell-style newtypes.

@@

* The statement to line 302 talks about moving values from the linear
  environment to the intuitionistic one. But only variables live in
  environments?

[TODO]

* I was confused by the comment on Linear Haskell starting on line
  610. What relevance does a system based on kinds have to NumLin
  (which isn't based on kinds either)?

We were trying to explain that (a) it seems like we can use GADTs with
natural-number indices to encode fractions a la Numlin in Linear
Haskell, but (b) we don't see how to support linear values (since
Linear Haskell only supports a linear arrow). We're happy to tighten
up the discussion in the final version.

@@

### Review C

* Compared to the learning curve of Rust (which the authors cite) what
  is the learning curve of Numlin, in the authors' estimation?

The basic intuitions about ownership are similar, so we expect an
experienced Rust programmer would pick up Numlin pretty
easily. However, Rust is a much bigger language, supporting a lot more
features than linear types, so the other direction would be more work!

@@

* Could the authors sketch the evolution of linear types from its
  initial classical introduction (say in Wadler's "Linear types can
  change the world" paper) to its current sophisticated form? Is this
  really an "off the shelf" linear type system?

We'll add this historical discussion to the final paper. Our system is
indeed fairly close to off-the-shelf, and as we mentioned to reviewer
A, it was one of the surprising discoveries of this paper that this
works.

@@

* Can they explain their soundness theorem once in English?

[TODO]

* My biggest comment is concerning concurrency. These sort of routines
  will be used in a setting where multicore concurrency will be
  employed to perform concurrent updates on arrays. In that setting,
  one might employ separation logic to deal with safe concurrent
  accesses. Now, separation logic also employs fractional
  permissions. Would Numlin be usable in that setting? Would the use
  of Separation Logic and Numlin (Linear Types) be orthogonal? Or do
  the authors feel the need to work out a concurrency extension of
  Numlin?

Numlin's type system enforces a concurrent-reader/exclusive-writer
pattern through the use of fractions, and our heap model was
deliberately chosen to be quite similar to the work on fractions from
separation logic. As a result, we would expect that proving that it
works in the presence of concurrency would be relatively
straightforward -- though without doing the work we don't want to say
that it would definitely work.

That said, developing a variant of separation logic suitable for
Numlin would be interesting, because it would let us prove things
about (eg) array bounds without having to throw them into the type
system.

### 2019-03-14 write-up/response.md

# Questions:

## Questions common to multiple reviews

* Reviews A and C both ask how we got the NumLin types for the
  BLAS/LAPACK routines.

The documentation records the expected aliasing behaviour of each of
these routines, and whether or not the routine modifies or consumes
its argument in any way. We used that to derive the types we ascribed
to the low-level routines. Since most of these low-level routines are
very careful not to do any allocation themselves, it was generally
very easy to give them a NumLin type -- every argument that can
modify/consume its argument needs a full permission, and all others
can be fraction-polymorphic. Taking Fortran as an example, it has
a notion of `in`, `out` and `inout` parameters. The latter two would
need full permissions, the first would be fraction-polymorphic.

@@

We certainly agree that both linear types and fractional permissions
have been studied before. To our surprise, it did not appear that
linear types had been much applied to checking linear algebra
libraries. It was only when we tried to use it, that we realized that
we needed fractional permissions.

In our view, this explains why it hadn't been used before: many prior
type systems for fractional permissions (eg, from Boyland's original
work to Aldrich and his students' work on Plural) are extremely
complex. This is because these type systems typically encode a complex
analysis to automatically infer how fractional permissions should be
split and rejoined.

By requiring sharing and merging to be made explicit, we were able to
drastically simplify the type system. Our formal system is very close
to standard presentations of linear logic, and the implementation
complexity is no worse than HM type inference.

* Finally, the related work focuses almost exclusively on type
  systems. Since ultimately the goal is to provide (memory) safe and
  fast linear algebra, I think it would be nice to also mention
  non-type based efforts, e.g. Spiral in Scala which compiles to
  efficient code [1], even though it does not use BLAS/LAPACK.  [1]
  Spiral in Scala: Towards the Systematic Construction of Generators
  for Performance Libraries. GPCE '13

We're happy to discuss this in the related work. However, NumLin is
more likely to be a consumer of efforts like Spiral than a competitor:
we want to check whether client code conforms to the aliasing rules
that low-level libraries (whether from platform libraries like BLAS or
generated via systems like Spiral, Cl1ck, FLAME) want clients to adhere to.

@@

### Review B

* One question I have is why OCaml code is generated? If the program
  only uses higher-order functions in certain ways, would it be
  possible to generate C code?

There's no technical obstacle to generating C code, but as we had
local access to the developer of the Owl numerical programming library
for OCaml, it made sense to use OCaml. Eventually we'd like to extend
Owl with something like NumLin to offer safe access to the low-level
platform numerics libraries.

@@

* Also, could the aliasing requirements be checked dynamically? I.e.,
  provide the OCaml interface in Appendix G, but with dynamic aliasing
  checks? What overhead would that have?

It might be possible, but we're not sure how. Our system enforces full
linearity (to support explicit deallocation), and to our knowledge
contract systems for linearity checking (eg, Tov and Pucella's ESOP
2010 "Stateful Contracts for Affine Types") only enforce affine usage
(ie, a dynamic check can enforce at-most-one, but not exactly-once).

* I was confused by the desugaring `let !x = e1 in e2` => `let Many x
  = e1 in let Many x = Many (Many x) in e2`. Can you explain this
  more?

There are two aspects to answering this, one is the type-based justification
and the second is *why* that's even necessary. Starting with why, the primary
motivation for having this was for binary arithmetic operations (of type
`!int * !int --o !int`). For the sake of ergonomics, we'd like to be able to write
code like "let x = 5 * 5 in x * x"; this is morally right because integers and
their operations are !-types but this is non-linear in x. Intuitively, we wish
for a rule like the following to "just do the right thing" when
destructuring/evaluating expressions of a !-type; which we can derive as a
meta-rule using the desugaring in question.

```
Int; Lin |- e : !t 
Int, x : !t; Lin' |- e' : t'
 ---------------------------------------
Int; Lin, Lin' |- let !x = e in e' : t'
```

@@

* I was initially confused by the presence (in Figure 11) of the
  seemling useless wrappers around banged values. Is this to make the
  API of the generated OCaml code easier to use correctly?

Yes, this is correct, we essentially embed the NumLin type system into the
OCaml one for a sanity check on the output code, ease of interop with OCaml and
tooling such as Merlin support.  Annotating with "[@@unboxed]" means OCaml's
underlying value representation does not allocate anything for unary datatypes,
so there is no performance penalty for using them like Haskell-style newtypes.

@@

* The statement to line 302 talks about moving values from the linear
  environment to the intuitionistic one. But only variables live in
  environments?

Yes, that's correct, this is a typo, we meant to write "moves the variable
referring to that value from the linear environment to the intuitionistic one".

@@

* I was confused by the comment on Linear Haskell starting on line
  610. What relevance does a system based on kinds have to NumLin
  (which isn't based on kinds either)?

We were trying to explain that (a) it seems like we can use GADTs with
natural-number indices to encode fractions a la NumLin in Linear
Haskell, but (b) we don't see how to support linear values (since
Linear Haskell only supports a linear arrow). We're happy to tighten
up the discussion in the final version.

@@

### Review C

* Compared to the learning curve of Rust (which the authors cite) what
  is the learning curve of NumLin, in the authors' estimation?

We conjecture that the learning curve of NumLin is less steep than that of
Rust, for two reaons: Rust supports more than just linear types and is smarter
about inferring the appropriate ownership. We suspect the explicit nature of
NumLin makes it easier to pick up and sets up an easier transition to Rust's
more complex system. We will add a mention of Weiss et. al. (2018/19) which
proposes a fraction-based framework to understand/model Rust's borrow checker.

@@

* Could the authors sketch the evolution of linear types from its
  initial classical introduction (say in Wadler's "Linear types can
  change the world" paper) to its current sophisticated form? Is this
  really an "off the shelf" linear type system?

We'll add this historical discussion to the final paper. Our system is
indeed fairly close to off-the-shelf, and as we mentioned to reviewer
A, it was one of the surprising discoveries of this paper that this
works.

* Can they explain their soundness theorem once in English?

We'll add an explanation of the following terse explanation to the final paper.
If an expression is syntactically type-checked, then for an arbitrary number of
steps (k), under *any* substitution of free fractional-permission variables,
linear variables (with a suitable heap) and intuitionistic variables, that
suitable heap and substituted expression are in the computational
interpretation (explained in English in the paper) of the (substituted) type of
that expression.

* My biggest comment is concerning concurrency. These sort of routines
  will be used in a setting where multicore concurrency will be
  employed to perform concurrent updates on arrays. In that setting,
  one might employ separation logic to deal with safe concurrent
  accesses. Now, separation logic also employs fractional
  permissions. Would NumLin be usable in that setting? Would the use
  of Separation Logic and NumLin (Linear Types) be orthogonal? Or do
  the authors feel the need to work out a concurrency extension of
  NumLin?

NumLin's type system enforces a concurrent-reader/exclusive-writer
pattern through the use of fractions, and our heap model was
deliberately chosen to be quite similar to the work on fractions from
separation logic. As a result, we would expect that proving that it
works in the presence of concurrency would be relatively
straightforward -- though without doing the work we don't want to say
that it would definitely work.

That said, developing a variant of separation logic suitable for
NumLin would be interesting, because it would let us prove things
about (eg) array bounds without having to throw them into the type
system.

### 2019-03-14 write-up/response.md

## Major Questions

* Reviews A and C both ask how we got the NumLin types for the
  BLAS/LAPACK routines.

    The documentation records the expected aliasing behaviour of each of
these routines, and whether or not the routine modifies or consumes
its argument in any way. We used that to derive the types we ascribed
to the low-level routines. Since most of these low-level routines are
very careful not to do any allocation themselves, it was generally
very easy to give them a NumLin type -- every argument that can
modify/consume its argument needs a full permission, and all others
can be fraction-polymorphic. Taking Fortran as an example, it has
a notion of `in`, `out` and `inout` parameters. The latter two would
need full permissions, the first would be fraction-polymorphic.

* Reviewer A asks:

    > The exact research contribution of this paper is, however, unclear
    > to me. Linear types as well as fractional permissions have been used
    > before.  Is the contibution that these two type systems are applied
    > together, or that they are applied to a new domain? If it is the
    > latter, it would be helpful if the paper could point out differences
    > to previous type systems, or challenges, if there are any.

    In our view, this explains why it hadn't been used before: many prior
type systems for fractional permissions (eg, from Boyland's original
work to Aldrich and his collaborators' work on Plural) are extremely
complex. This is because these type systems typically encode a complex
analysis to automatically infer how fractional permissions should be
split and rejoined.

* Reviewer C asks:

    > My biggest comment is concerning concurrency. These sort of routines
    > will be used in a setting where multicore concurrency will be
    > employed to perform concurrent updates on arrays. In that setting,
    > one might employ separation logic to deal with safe concurrent
    > accesses. Now, separation logic also employs fractional
    > permissions. Would NumLin be usable in that setting? Would the use
    > of Separation Logic and NumLin (Linear Types) be orthogonal? Or do
    > the authors feel the need to work out a concurrency extension of
    > NumLin?

    NumLin's type system enforces a concurrent-reader/exclusive-writer
pattern through the use of fractions (a pattern widely used in
concurrent programming), and our heap model was deliberately chosen to
be quite similar to the work on fractions from separation logic. As a
result, we would expect that the language would work unchanged in the
concurrent setting, and that extending the proof to show that it works
in the presence of concurrency would be relatively straightforward --
though without doing the work we don't want to say for sure that it
would definitely work.

----

@@

----

## Reviewer-specific questions

### Review A

* > Finally, the related work focuses almost exclusively on type
  > systems. Since ultimately the goal is to provide (memory) safe and
  > fast linear algebra, I think it would be nice to also mention
  > non-type based efforts, e.g. Spiral in Scala which compiles to
  > efficient code [1], even though it does not use BLAS/LAPACK.  [1]
  > Spiral in Scala: Towards the Systematic Construction of Generators
  > for Performance Libraries. GPCE '13

@@

### Review B

* > One question I have is why OCaml code is generated? If the program
  only uses higher-order functions in certain ways, would it be
  possible to generate C code?

    There's no technical obstacle to generating C code, but as we had
local access to the developer of the Owl numerical programming library
for OCaml, it made sense to use OCaml. Eventually we'd like to extend
Owl with something like NumLin to offer safe access to the low-level
platform numerics libraries.

* > Also, could the aliasing requirements be checked dynamically? I.e.,
  provide the OCaml interface in Appendix G, but with dynamic aliasing
  checks? What overhead would that have?

@@

    It might be possible, but we're not sure how. Our system enforces full
linearity (to support explicit deallocation), and to our knowledge
contract systems for linearity checking (eg, Tov and Pucella's ESOP
2010 "Stateful Contracts for Affine Types") only enforce affine usage
(ie, a dynamic check can enforce at-most-one, but not exactly-once).

* > I was confused by the desugaring `let !x = e1 in e2` => `let Many x
  = e1 in let Many x = Many (Many x) in e2`. Can you explain this
  more?

    There are two aspects to answering this, one is the type-based justification
and the second is *why* that's even necessary. Starting with why, the primary
motivation for having this was for binary arithmetic operations (of type
`!int * !int --o !int`). For the sake of ergonomics, we'd like to be able to write
code like "let x = 5 + 5 in x - x"; this is morally right because integers and
their operations are !-types but this is non-linear in x. Intuitively, we wish
for a rule like the following to "just do the right thing" when
destructuring/evaluating expressions of a !-type; which we can derive as a
meta-rule using the desugaring in question.

  > ```
Int; Lin |- e : !t 
Int, x : !t; Lin' |- e' : t'
 ---------------------------------------
Int; Lin, Lin' |- let !x = e in e' : t'
```

   It should be possible to handle this using the Benton/Wadler
LNL-style presentation of linear types, which uses adjoint modalities
to distinguish between intrinsically linear and intrinsically
nonlinear types, but that is a pretty big digression from our main
goal.

* > I was initially confused by the presence (in Figure 11) of the
  seemling useless wrappers around banged values. Is this to make the
  API of the generated OCaml code easier to use correctly?

@@

    Yes, this is correct, we essentially embed the NumLin type system into the
OCaml one for a sanity check on the output code, ease of interop with OCaml and
tooling such as Merlin support.  Annotating with "[@@unboxed]" means OCaml's
underlying value representation does not allocate anything for unary datatypes,
so there is no performance penalty for using them like Haskell-style newtypes.

* > The statement to line 302 talks about moving values from the linear
  environment to the intuitionistic one. But only variables live in
  environments?

@@

    Yes, that's correct, this is a typo, we meant to write "moves the variable
referring to that value from the linear environment to the intuitionistic one".

* > I was confused by the comment on Linear Haskell starting on line 610. 
  What relevance does a system based on kinds have to NumLin
  (which isn't based on kinds either)?

@@

    We were trying to explain that (a) it seems like we can use GADTs with
natural-number indices to encode fractions a la NumLin in Linear
Haskell, but (b) we don't see how to support linear values (since
Linear Haskell only supports a linear arrow). We're happy to tighten
up the discussion in the final version.

### Review C

* > Compared to the learning curve of Rust (which the authors cite) what
  is the learning curve of NumLin, in the authors' estimation?

@@

    We conjecture that the learning curve of NumLin is less steep than that of
Rust, for two reaons: Rust supports more than just linear types and is smarter
about inferring the appropriate ownership. We suspect the explicit nature of
NumLin makes it easier to pick up and sets up an easier transition to Rust's
more complex system. We will add a mention of Weiss et. al. (2018/19) which
proposes a fraction-based framework to understand/model Rust's borrow checker.

* > Could the authors sketch the evolution of linear types from its
  initial classical introduction (say in Wadler's "Linear types can
  change the world" paper) to its current sophisticated form? Is this
  really an "off the shelf" linear type system?

@@

    We'll add this historical discussion to the final paper. Our system is
indeed fairly close to off-the-shelf, and as we mentioned to reviewer
A, it was one of the surprising discoveries of this paper that this
works.

* > Can they explain their soundness theorem once in English?

    We'll add an expansion of the following terse explanation to the
final paper. If an expression is syntactically type-checked, then for
an arbitrary number of steps (k), under *any* substitution of free
fractional-permission variables, linear variables (with a suitable
heap) and intuitionistic variables, that suitable heap and substituted
expression are in the computational interpretation (explained in
English in the paper) of the (substituted) type of that expression.

    The "computational interpretation" just identifies executions that do
no un- or ill-defined behaviours (eg, adding a boolean or an integer).
Since our operational semantics explicitly models deallocation, we now
know no well-typed program will ever try to access deallocated memory,
establishing the correctness of our memory management checking.

### 2019-03-14 write-up/response.md

* > Also, could the aliasing requirements be checked dynamically? I.e.,
  provide the OCaml interface in Appendix G, but with dynamic aliasing
  checks? What overhead would that have?

    It might be possible, but we're not sure how. Our system enforces full
linearity (to support explicit deallocation), and to our knowledge
contract systems for linearity checking (eg, Tov and Pucella's ESOP
2010 "Stateful Contracts for Affine Types") only enforce affine usage
(ie, a dynamic check can enforce at-most-one, but not exactly-once).

* > I was confused by the desugaring `let !x = e1 in e2` => `let Many x
  = e1 in let Many x = Many (Many x) in e2`. Can you explain this
  more?

    There are two aspects to answering this, one is the type-based justification
and the second is *why* that's even necessary. Starting with why, the primary
motivation for having this was for binary arithmetic operations (of type
`!int * !int --o !int`). For the sake of ergonomics, we'd like to be able to write
code like `let x = 5 + 5 in x - x`; this is morally right because integers and
their operations are !-types but this is non-linear in x. Intuitively, we wish
for a rule like the following to "just do the right thing" when
destructuring/evaluating expressions of a !-type; which we can derive as a
meta-rule using the desugaring in question.

    ```
    Int; Lin |- e : !t
    Int, x : !t; Lin' |- e' : t'
    ---------------------------------------
    Int; Lin, Lin' |- let !x = e in e' : t'
    ```

@@

* > Could the authors sketch the evolution of linear types from its
  initial classical introduction (say in Wadler's "Linear types can
  change the world" paper) to its current sophisticated form? Is this
  really an "off the shelf" linear type system?

    We'll add this historical discussion to the final paper. Our system is
indeed fairly close to off-the-shelf, and as we mentioned to reviewer
A, it was one of the surprising discoveries of this paper that this
works.

* > Can they explain their soundness theorem once in English?

    We'll add an expansion of the following terse explanation to the
final paper. If an expression is syntactically type-checked, then for
an arbitrary number of steps (k), under *any* substitution of free
fractional-permission variables, linear variables (with a suitable
heap) and intuitionistic variables, that suitable heap and expression
are in the computational interpretation of the type of that expression.

### 2019-04-09 write-up/paper/intro.tex
Fix for shepherding: improve introduction
to a) emphasise that the main contribution of this paper is the
definition of the NumLin language and its application as a safe language
for invoking BLAS routines, and b) to clarify the relationship to
previous work on linear types and fractional permissions (as explained
in the response).

\lang\ is a functional programming language designed to express the
APIs of low-level linear algebra libraries (such as BLAS/LAPACK)
safely and explicitly.  It does so by combining linear types,
fractional permissions, runtime errors and recursion into a small,
easily understandable, yet expressive set of core constructs.

>>>>

\lang\ is a functional programming language whose type system is
designed to enforce the safe usage of the APIs of low-level linear algebra
libraries (such as BLAS/LAPACK).  It does so by combining linear types,
fractional permissions, runtime errors and recursion into a small, easily
understandable, yet expressive set of core constructs.

### 2019-04-09 write-up/paper/paper.tex
Add title
Beefed up intro; added a stronger opening to language overview
Fix for shepherding: improve introduction
to a) emphasise that the main contribution of this paper is the
definition of the NumLin language and its application as a safe language
for invoking BLAS routines, and b) to clarify the relationship to
previous work on linear types and fractional permissions (as explained
in the response).

\begin{abstract}
    We present \lang, a functional programming language designed to express
    the APIs of low-level linear algebra libraries (such as BLAS/LAPACK) safely
    and explicitly, through a brief description of its key features and several
    illustrative examples. We show that \lang's type system is sound and that
    its implementation improves upon na{\"i}ve implementations of linear
    algebra programs, almost towards C-levels of performance. Lastly, we contrast
    it to other recent developments in linear types and show that using linear
    types and fractional permissions to express the APIs of low-level linear
    algebra libraries is a simple and effective idea.
\end{abstract}

>>>>

\maketitle%

\begin{abstract}
    We present \lang, a functional programming language whose type system is
    designed to enforce the safe usage of the APIs of low-level linear algebra
    libraries (such as BLAS/LAPACK). We do so through a brief description of
    its key features and several illustrative examples. We show that \lang's
    type system is sound and that its implementation improves upon na{\"i}ve
    implementations of linear algebra programs, almost towards C-levels of
    performance. By doing so, we demonstrate (a) that linear types are
    well-suited to expressing the APIs of low-level linear algebra libraries
    accurately and concisely and (b) that, despite the complexity of prior work
    on it, fractional permissions can actually be implemented using simple,
    well-known techniques and be used practically in real programs.
\end{abstract}
